### 2019-03-18 response/response-demography-1.tex

\section*{Editor}

First, the description of the problem and literature review are sparse.
The front end of the paper should be revised to situate the study in the
wider literature on son preference and sex selection in the region. The
contributions of the current study should be highlighted. R1 offered
particularly useful ideas in this regard. These changes would make the
paper appeal to the broad readership of Demography.

[Response:]

I have completely rewritten the introduction to better situate the paper in the literature
and to highlight the contributions of the paper.
I now cover the research on using spacing as a measure of son preference, the 
relationship between spacing and child health (both older and younger child), the 
effect on women's health and other outcomes, and, finally, the limited and mixed results
on what determines birth spacing behavior in developing countries compared to developed 
countries.
I use these to motivate why examining how birth spacing has been affected by the use of
sex selection is important, and, in addition, highlight the three main questions I address 
in the paper: the effect of sex selection on birth intervals, the effect of sex selection
on the distribution of births within spells, and, finally, how the effects have varied
with geographical region.

Second, a conceptual framework or theoretical argument is needed to
motivate the decision to examine birth intervals and sex composition by
maternal education. Does maternal education proxy for access to sex
selective technology, preferences for sons, etc.?

In addition, how does
selection into the highest/lowest educated categories change over the 45
years of the study? How could this impact the interpretation of results?

@@

The cultural explanation also suggests a reason for why the states son preference 
appear to decline over time.
The ``newer'' entrants have lower initial son preference and therefore the average
son preference for high education women will decline over time.
\citep{bhat03}, \citep{pande07}.

Third, the paper is heavy on the description of methods and results.
These sections could be shortened and reorganized to focus on the most
important issues. R1 also offered suggestions regarding the presentation
of findings.

[Response:]

As detailed below, I have substantially shorten the ``Estimation Strategy'' section.
The results section is not necessarily shorter, but is better focused on the spacing
results, guided by the three main questions.
There is also less focus on the sex selection results.
Specifically, I have removed the section on how the use of sex-selective abortions changes 
within a spell since it, as originally written, was more focused on sex selection than
the spacing, and that the spacing question is better addressed in the section on the
distribution of births within spells.
Obviously I still need to discuss how sex selection has changed to tie those to the 
changes in spacing, but I have tried to make these result mainly serve the overall
spacing analysis.

### 2019-03-19 response/response-demography-1.tex

\begin{center} Response to Editor and Referee Comments \end{center}

\noindent Please find attached the revised version of my paper,
``Birth Spacing in the Presence of Son Preference and Sex-Selective
Abortions:
India's Experience over Four Decades.''
I would like to thank the two referees and the editor for the very useful comments and 
suggestions, which I believe substantially improved the paper.

\section*{Editor}

\begin{enumerate}

\item First, the description of the problem and literature review are sparse.
The front end of the paper should be revised to situate the study in the
wider literature on son preference and sex selection in the region. The
contributions of the current study should be highlighted. R1 offered
particularly useful ideas in this regard. These changes would make the
paper appeal to the broad readership of Demography.

@@

[Response:]

I have completely rewritten the introduction to better situate the paper in the literature
and to highlight the contributions of the paper.
I now cover the research on using spacing as a measure of son preference, the 
relationship between spacing and child health (both older and younger child), the 
effect on women's health and other outcomes, and, finally, the limited and mixed results
on what determines birth spacing behavior in developing countries compared to developed 
countries.
I use these to motivate why examining how birth spacing has been affected by the use of
sex selection is important, and, in addition, highlight the three main questions I address 
in the paper: the effect of sex selection on birth intervals, the effect of sex selection
on the distribution of births within spells, and, finally, how the effects have varied
with geographical region.

\item Second, a conceptual framework or theoretical argument is needed to
motivate the decision to examine birth intervals and sex composition by
maternal education. Does maternal education proxy for access to sex
selective technology, preferences for sons, etc.? 
In addition, how does
selection into the highest/lowest educated categories change over the 45
years of the study? How could this impact the interpretation of results?

@@

For the economics one, the use of sex selection is mainly driven by the lower fertility
that comes from higher opportunity cost of time and the higher income that allow those
with higher education to use sex selection.
Here there is essentially no selection bias since the use of sex selection is driven
by factors that are identical across those with the same level of education.

The cultural explanation is that originally those with higher education where also those
groups with the strongest son preference (presumably those who are also higher caste).
The expansion of access to education brought other groups/castes into connection with
the higher castes and assimilation of their son preference norms.
Presumably the assimilation is not complete and those newer to higher education would
therefore have lower son preference and therefore lower use of sex selection.
This means that any appearance of lower use of sex selection could be a selection story,
and that the initial group's son preference is just as strong.

The cultural explanation also suggests a reason for why the states son preference 
appear to decline over time.
The ``newer'' entrants have lower initial son preference and therefore the average
son preference for high education women will decline over time.
\citep{bhat03}, \citep{pande07}.

\item Third, the paper is heavy on the description of methods and results.
These sections could be shortened and reorganized to focus on the most
important issues. R1 also offered suggestions regarding the presentation
of findings.

@@

[Response:]

As detailed below, I have substantially shorten the ``Estimation Strategy'' section.
The results section is not necessarily shorter, but is better focused on the spacing
results, guided by the three main questions.
There is also less focus on the sex selection results.
Specifically, I have removed the section on how the use of sex-selective abortions changes 
within a spell since it, as originally written, was more focused on sex selection than
the spacing, and that the spacing question is better addressed in the section on the
distribution of births within spells.
Obviously I still need to discuss how sex selection has changed to tie those to the 
changes in spacing, but I have tried to make these result mainly serve the overall
spacing analysis.

\item Fourth, discussion of the fertility transition is neglected and only
discussed in the conclusion. At a minimum, you should consider providing
details on fertility levels and distributions across parities over time.
Is there an easy way to show the sample sizes/proportion of births in
each category of sex composition across years in tables 1-3? More could
be said about the potential impact of the ongoing fertility decline.

@@

[Response:]

I address this a number of different ways.
First, in the Introduction, I discuss the decline in fertility when introducing why India 
is an interesting case, with special focus on the relationship between declining fertility 
and increased use of sex selection and, hence, by my hypothesis on birth spacing.
Second, the suggestion by Referee 1 to use graphical presentation of the results made it 
possible to include the likelihood of parity progression for each of the three spells, 
which allows me to discuss how fertility has decreased across the three education groups
and by urban and rural areas.
I have also added the parity progression likelihood to the original tables, which are now
included in the online Appendix.
Third, I discuss the changes in fertility directly in the Results section.

% [Generically, spacing tend to decline with lower fertility (check this). 
% Presumably the decline in spacing is because part of the cause is higher female labor 
% market participation, which makes it more attractive to space children closer together.
% Not clear whether this is the case for India - The R1 suggested paper might be a good
% starting point with a tie-in to compression.
% Compression can also come from improved health and/or less migration.
% 
% On the flip side, lower fertility place additional pressure on women to use
% sex selection, which lead to *less* compression.
% ]

\end{enumerate}

### 2019-03-19 paper/sex_selection_spacing-ver2.tex
Edit Introduction

Birth spacing has long served as a measure of son preference, with strong
son preference typically associated with shorter birth intervals the
fewer sons a family has \citep{ben-porath76b,Leung1988}.
However, the introduction of prenatal sex determination has the potential to
fundamentally change the relationship between son preference and birth spacing.
The stronger the son preference, the more likely couples are to resort to sex-selective 
abortions, and each abortion automatically increases the interval between births by six 
months to a year.%
\footnote{
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Longer birth intervals may, therefore, cease to be an indication of declining son
preference, but could instead come about precisely because strong son preference makes
couples more likely to use sex selection.

@@

% [Why India?]
India is a particularly compelling case.
India has a long history of male preference, especially in the northern states 
\citep{Kishor1993,murthi95,arnold98}.
A significant proportion of Indian families exhibit a son preference based 
differential stopping behavior, where, for a given number of children, they are more 
likely to have an additional child if they have not yet reached their preferred number of 
sons \citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.%
\footnote{
Other countries show a similar pattern
\citep[see, for example,][]{larsen98,filmer09,Altindag2016}.
}
Moreover, mortality risk is higher for females than males for most ages groups, and
there has been an almost continuous increase in India's overall ratio of males to females 
over the last century \citep{dyson01,Navaneetham2011,Bongaarts2015}.

% [What do we know about use of sex selection?]
Over the last three decades, India has seen dramatic increases in the males-to-females 
ratio at birth as access to prenatal sex determination expanded 
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06,Guilmoto2012,Portner2015b}.
Simultaneously, fertility has declined substantially, to the point where it is now at, or 
even below, replacement in some areas 
\citep{Guilmoto2013,Dharmalingam2014,International-Institute-for-Population-Sciences-IIPS2017}.
This decline in fertility, combined with a continued strong preference for sons, has 
intensified the pressure to use sex selection 
\citep{das_gupta97,Guilmoto2009a,Bongaarts2013,Jayachandran2017}.%
\footnote{
In some instances, however, the fertility decline appears to have created a stronger 
aversion to daughters rather than a stronger preference for son \citet{DiamondSmith2008}.
}
At household level the strongest predictor of the use of sex selection is the sex 
composition of previous children \citep{retherford03b,jha06,abrevaya09,Portner2015b}.
For families without a son, the higher the parity, the higher is the probability
of having a son as the next birth.
The use of sex selection also increases with socioeconomic status, especially 
education, and is more prevalent in cities than in rural areas, both possibly because 
of lower desired fertility \citep{retherford03b,jha06,Portner2015b}.

@@

Even the effects of access to contraceptive and declining fertility on spacing are unclear.
On the one hand, access to contraceptives allows women to avoid too short spacing between 
birth, which would increase birth spacing.
On the other hand, increased reliability of access and effectiveness of contraceptives can 
lead to shorter intervals between births  \citep{Keyfitz1971,Heckman1976}.
With less reliable contraception parents choose a higher level of contraception, which 
results in longer spacing, to avoid having too many children by accident.
However, as contraception becomes more effective parents can more easily avoid future 
births, allowing them to reduce the spacing between births without having to worry about 
overshooting their preferred number of children.
This idea may explain why better-educated women have shorter spacing than less educated 
women in some instances but not in others.
These two counteracting effects may explain why finding statistically significant effects 
of contraception use on birth spacing is difficult \citep{Yeakey2009}.%
\footnote{
There is evidence in Matlab that the effect of son preference on birth spacing is
stronger in areas with better access to family planning \citep{Rahman1993}.
Other evidence, however, points to families ability to time birth, even in the absence
of modern contraceptives \citep{Jayachandran2011,Alam2018}.
}

@@

Second, have the shapes of the survival curves changed?
The shapes of the survival curves provide information both on general changes in birth 
spacing and whether parents are still trying to conceive earlier after girls than boys.
I still expect parents to \emph{try} to conceive earlier after the birth 
of a girl than after a boy, but the abortions will increase the duration between births, 
and may even lead to longer average spacing after the birth of girls than after boys.
I expect this to be the case if there is still a strong preference for boys, and this 
behavior may even become more pronounced with access to sex selection.

@@

% [Results - need updating]
There are three main results.
First, there has been a general increase in the length of spacing between births
over the four decades covered by the data.
Second, the most substantial increase in spacing is for the women who
are most likely to use sex selection.
Specifically, among the best-educated women, those with no boys now has 
significantly longer spacing---and a more male-biased sex ratio---than 
similar women with boys.
Third, women with no education still follow the standard pattern of
short spacing when they have girls and little evidence of the use of sex
selection.
In other words, these women adhere to a strategy where they achieve a
son through higher fertility rather than the use of sex selection.

>>>>

Birth spacing has long served as a measure of son preference, with strong
son preference typically associated with shorter birth intervals the
fewer sons a family has \citep{ben-porath76b,Leung1988}.
However, the introduction of prenatal sex determination has the potential to
fundamentally change the relationship between son preference and birth spacing.
The stronger the son preference, the more likely couples are to resort to sex-selective 
abortions, and each abortion automatically increases the interval between births by six 
months to a year.%
\footnote{
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Longer birth intervals in the absence of sons, which previously would be taken to indicate 
a lower son preference, now may arise precisely because a sufficiently strong son 
preference makes couples use sex selection.

@@

% [Why India?]
India is a particularly compelling case since it has a long history of male preference, 
especially in the northern states \citep{Kishor1993,murthi95,arnold98}.
A significant proportion of Indian families exhibit differential stopping behavior, where, 
for a given number of children, they are more likely to have an additional child if they 
have not yet reached their preferred number of sons 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.%
\footnote{
Other countries show a similar pattern
\citep[see, for example,][]{larsen98,filmer09,Altindag2016}.
}
Moreover, mortality risk is higher for females than males for most ages groups, and
there has been an almost continuous increase in India's overall ratio of males to females 
over the last century \citep{dyson01,Navaneetham2011,Bongaarts2015}.

% [What do we know about use of sex selection?]
Over the last three decades, India has seen dramatic increases in the males-to-females 
ratio at birth as access to prenatal sex determination expanded 
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06,Guilmoto2012,Portner2015b}.
Simultaneously, fertility has declined substantially, to the point where it is now at, or 
even below, replacement in some areas 
\citep{Guilmoto2013,Dharmalingam2014,International-Institute-for-Population-Sciences-IIPS2017}.
This decline in fertility, combined with a continued strong preference for sons, has 
intensified the pressure to use sex selection 
\citep{das_gupta97,Guilmoto2009a,Bongaarts2013,Jayachandran2017}.%
\footnote{
In some instances, however, the fertility decline appears to have created a stronger 
aversion to daughters rather than a stronger preference for son \citep{DiamondSmith2008}.
}
The strongest predictor of the use of sex selection is the sex composition of previous 
children; for families without a son, the higher the parity, the higher is the probability
of having a son as the next birth \citep{retherford03b,jha06,abrevaya09,Portner2015b}.
The use of sex selection also increases with socioeconomic status, especially 
education, and is more prevalent in cities than in rural areas, both possibly because 
of lower desired fertility \citep{retherford03b,jha06,Portner2015b}.

@@

Even the effects of access to contraceptive and declining fertility on spacing are unclear.
On the one hand, access to contraceptives allows women to avoid too short spacing between 
birth, which would increase birth spacing.
On the other hand, increased reliability of access and effectiveness of contraceptives can 
lead to shorter intervals between births  \citep{Keyfitz1971,Heckman1976}.
With less reliable contraception parents choose a higher level of contraception, which 
results in longer spacing, to avoid having too many children by accident.
However, as contraception becomes more effective parents can more easily avoid future 
births, allowing them to reduce the spacing between births without having to worry about 
overshooting their preferred number of children.
These two counteracting effects may explain why better-educated women have shorter spacing 
than less educated women in some instances but not in others and why finding statistically 
significant effects of contraception use on birth spacing is difficult \citep{Yeakey2009}.%
\footnote{
There is evidence in Matlab that the effect of son preference on birth spacing is
stronger in areas with better access to family planning \citep{Rahman1993}.
Other evidence, however, points to families ability to time birth, even in the absence
of modern contraceptives \citep{Jayachandran2011,Alam2018}.
}

@@

[TK change based on rewrite of section]
Second, have the shapes of the survival curves changed?
The shapes of the survival curves provide information both on general changes in birth 
spacing and whether parents are still trying to conceive earlier after girls than boys.
I still expect parents to \emph{try} to conceive earlier after the birth 
of a girl than after a boy, but the abortions will increase the duration between births, 
and may even lead to longer average spacing after the birth of girls than after boys.
I expect this to be the case if there is still a strong preference for boys, and this 
behavior may even become more pronounced with access to sex selection.

@@

% [TK Results - need updating]
There are three main results.
First, there has been a general increase in the length of spacing between births
over the four decades covered by the data.
Second, the most substantial increase in spacing is for the women who
are most likely to use sex selection.
Specifically, among the best-educated women, those with no boys now has 
significantly longer spacing---and a more male-biased sex ratio---than 
similar women with boys.
Third, women with no education still follow the standard pattern of
short spacing when they have girls and little evidence of the use of sex
selection.
In other words, these women adhere to a strategy where they achieve a
son through higher fertility rather than the use of sex selection.
